---
title: "Getting started with clinAnnotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with clinAnnotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 8,
  dpi       = 150,
  out.width = "100%"
)
```

clinAnnotR generates multi-panel figures that combine numerical laboratory
time-series with per-case treatment Gantt charts. Any parameter can be shown as
a line or point series on log10 or linear axes, with below-detection-limit (BDL)
handling and optional direct value labels. This vignette walks through the
complete workflow from loading data to saving a publication-ready figure.

```{r load-package}
library(clinAnnotR)
```

---

## 1  Data formats

clinAnnotR expects two data frames.

### Lab values (long format)

| Column      | Type      | Description                                      |
|-------------|-----------|--------------------------------------------------|
| `case_id`   | character | Unique case identifier                           |
| `relday`    | numeric   | Days from the reference date (day 0)             |
| `parameter` | character | Measurement label, e.g. `"WBC (/µL)"`           |
| `value`     | character | Raw value; prefix `"<X"` flags a BDL result      |

### Treatment records (wide-ish format)

| Column      | Type      | Description                                              |
|-------------|-----------|----------------------------------------------------------|
| `case_id`   | character | Must match `lab_data$case_id`                            |
| `TREATMENT` | character | Drug or intervention name                                |
| `START_rel` | numeric   | Treatment start (relative days)                          |
| `END_rel`   | numeric   | Treatment end (relative days)                            |
| `COLOR`     | character | Hex fill colour for the Gantt bar, e.g. `"#88C0D0"`     |
| `CLASS`     | character | Drug class label, e.g. `"Antimetabolite"`                |

Multiple rows with the same `TREATMENT` for one case are rendered as a
multi-segment Gantt bar, which is useful for cyclic regimens.

---

## 2  Loading data

### Option A — from a CSV or data frame

Use `prep_lab_data()` and `prep_treatment_data()` when relative days are already
computed (e.g. data read from a CSV, or assembled inline):

```{r prep-from-df}
raw_lab <- data.frame(
  case_id   = c("Case 1", "Case 1", "Case 1", "Case 2", "Case 2"),
  relday    = c(0, 22, 49, 0, 22),
  parameter = c("WBC (/µL)", "WBC (/µL)", "WBC (/µL)",
                "WBC (/µL)", "WBC (/µL)"),
  value     = c("48500", "4400", "2800", "27300", "3200")
)

raw_tx <- data.frame(
  case_id   = c("Case 1", "Case 1", "Case 2"),
  TREATMENT = c("Cytarabine", "Venetoclax", "Cytarabine"),
  START_rel = c(1, 49, 1),
  END_rel   = c(7, 185, 7),
  COLOR     = c("#88C0D0", "#B48EAD", "#88C0D0"),
  CLASS     = c("Antimetabolite", "BCL2 inhibitor", "Antimetabolite")
)

lab <- prep_lab_data(raw_lab)
tx  <- prep_treatment_data(raw_tx)
```

### Option B — from an Excel file

Use `load_lab_data()` and `load_treatment_data()` when your source data contains
absolute dates. Pass a `cases` list that pairs each case identifier with its
reference date (day 0):

```{r load-excel, eval = FALSE}
cases <- list(
  list(id = "Case 1", ref_date = "2025-06-03"),
  list(id = "Case 2", ref_date = "2025-08-12")
)

lab <- load_lab_data("labvals.xlsx", cases)
tx  <- load_treatment_data("treatment.xlsx", cases,
                           col_patientid = "PATIENTID")
```

Both functions return the same standardised column layout as their
`prep_*` counterparts.

### Option C — built-in example data

A pair of fictional datasets covering two cases is included in the package:

```{r example-data}
lab <- example_data("lab")
tx  <- example_data("treatment")

head(lab, 4)
head(tx, 4)
```

---

## 3  Defining lab panels

Each time-series panel is described by a `lab_panel_spec` object created with
`lab_panel()`. You can have any number of panels; they are stacked vertically
in the order you supply them.

A spec has no ggplot code — it is just a list of instructions that
`make_clinical_figure()` interprets later.

```{r panel-specs}
# Panel 1: WBC line + peripheral blast points on a shared log10 count axis
panel_counts <- lab_panel(
  line_params  = "WBC (/µL)",
  point_params = "peripheral blasts (/µL)",
  y_label      = "Count (/µL)",
  height_weight = 4
)

# Panel 2: BM blast percentages with direct value labels
panel_blasts <- lab_panel(
  point_params  = "BM blasts with IF (%)",
  show_labels   = TRUE,
  label_suffix  = "%",
  y_limits      = c(0.05, 200),
  y_breaks      = c(0.1, 1, 5, 20, 100),
  y_labels      = paste0(c(0.1, 1, 5, 20, 100), "%"),
  y_label       = "BM blasts (%)",
  height_weight = 3
)

print(panel_counts)
```

Key arguments:

| Argument       | Default   | Purpose                                               |
|----------------|-----------|-------------------------------------------------------|
| `line_params`  | `c()`     | Parameters drawn as connected lines                   |
| `point_params` | `c()`     | Parameters drawn as points (BDL uses a dedicated symbol) |
| `y_scale`      | `"log10"` | `"log10"` or `"linear"`                               |
| `y_label`      | auto      | y-axis title; `NULL` joins all parameter names        |
| `y_limits`     | auto      | `c(lo, hi)` for fixed axis limits                     |
| `y_breaks`     | auto      | Break positions on the y-axis                         |
| `y_labels`     | auto      | Tick labels (character vector or labelling function)   |
| `show_labels`  | `FALSE`   | Annotate each non-BDL point with its numeric value    |
| `label_suffix` | `""`      | Appended to every value label, e.g. `"%"`             |
| `height_weight`| `3`       | Relative panel height in the assembled figure          |
| `highlight_days`| `NULL`   | Per-panel reference lines (overrides global setting)  |

---

## 4  Assembling the figure

`make_clinical_figure()` combines all lab panels and the Gantt chart(s) into a
single patchwork object.

```{r make-figure}
fig <- make_clinical_figure(
  lab_data       = lab,
  treatment_data = tx,
  lab_panels     = list(panel_counts, panel_blasts),
  highlight_days = c("D1" = 1, "D22" = 22, "D49" = 49)
)

print(fig)
```

All panels share the same x-axis range. The x-axis label and ticks appear only
on the bottom-most panel. The combined legend (cases, parameters, BDL symbol)
is collected from the first lab panel and shown once.

### Key arguments

| Argument              | Default | Purpose                                                    |
|-----------------------|---------|------------------------------------------------------------|
| `lab_data`            | —       | Output of `prep_lab_data()` or `load_lab_data()`          |
| `treatment_data`      | `NULL`  | Output of `prep_treatment_data()` or `load_treatment_data()`; `NULL` omits Gantt panels |
| `lab_panels`          | —       | List of `lab_panel_spec` objects                           |
| `cases`               | auto    | Case IDs to include and their stacking order               |
| `case_palette`        | auto    | Named vector of hex colours per case                       |
| `x_range`             | auto    | `c(lo, hi)` in days; defaults to data range ± 5 days      |
| `highlight_days`      | `NULL`  | Named numeric vector of reference line positions           |
| `gantt_height_weight` | auto    | Fixed Gantt height; `NULL` scales with treatment row count |
| `shade_regions`       | `NULL`  | Background shading (see below)                             |
| `caption`             | auto    | Figure caption text                                        |
| `base_size`           | `9`     | Base font size applied to all panels                       |

---

## 5  Customisation

### Reference lines (`highlight_days`)

Pass a named numeric vector. Names appear as rotated labels at the top of the
first lab panel.

```{r highlight, eval = FALSE}
make_clinical_figure(
  lab_data       = lab,
  treatment_data = tx,
  lab_panels     = list(panel_counts),
  highlight_days = c("D1" = 1, "D22" = 22, "D49" = 49, "D100" = 100)
)
```

Individual panels can override the global `highlight_days` by supplying
`highlight_days` inside `lab_panel()`.

### Background shading (`shade_regions`)

Shade rectangular regions per case to highlight treatment phases or events.
Each region is a list with `xmin`, `xmax`, `fill`, and optional `alpha`
(default `0.15`):

```{r shade, eval = FALSE}
make_clinical_figure(
  lab_data       = lab,
  treatment_data = tx,
  lab_panels     = list(panel_counts, panel_blasts),
  highlight_days = c("D1" = 1, "D22" = 22, "D49" = 49),
  shade_regions  = list(
    "Case 1" = list(
      list(xmin = 1,  xmax = 7,   fill = "#88C0D0", alpha = 0.12),
      list(xmin = 49, xmax = 185, fill = "#B48EAD", alpha = 0.12)
    ),
    "Case 2" = list(
      list(xmin = 1,  xmax = 7,   fill = "#88C0D0", alpha = 0.12),
      list(xmin = 49, xmax = 185, fill = "#B48EAD", alpha = 0.12)
    )
  )
)
```

Shaded regions appear behind all data layers on every lab panel and on the
corresponding case's Gantt panel.

### Fixed x-axis range

```{r x-range, eval = FALSE}
make_clinical_figure(
  lab_data       = lab,
  treatment_data = tx,
  lab_panels     = list(panel_counts),
  x_range        = c(-10, 200)
)
```

### Custom case colours

```{r case-palette, eval = FALSE}
make_clinical_figure(
  lab_data      = lab,
  treatment_data = tx,
  lab_panels    = list(panel_counts),
  case_palette  = c("Case 1" = "#5E81AC", "Case 2" = "#BF616A")
)
```

### Linear-scale panel

```{r linear-panel, eval = FALSE}
panel_linear <- lab_panel(
  line_params = c("Hemoglobin (g/dL)", "Albumin (g/dL)"),
  y_scale     = "linear",
  y_label     = "Concentration (g/dL)"
)
```

### Single-case figures

When `lab_data` contains only one case, the colour mapping switches from
case to parameter automatically — all lines are solid and each parameter
gets a distinct colour. The Gantt panel is labelled with that case's ID.

```{r single-case}
lab1 <- lab[lab$case_id == "Case 1", ]
tx1  <- tx[tx$case_id  == "Case 1", ]

fig1 <- make_clinical_figure(
  lab_data       = lab1,
  treatment_data = tx1,
  lab_panels     = list(panel_counts, panel_blasts),
  highlight_days = c("D1" = 1, "D22" = 22, "D49" = 49)
)

print(fig1)
```

---

## 6  Below-detection-limit (BDL) values

Values prefixed with `"<"` in the raw data (e.g. `"<100"`) are flagged as BDL
and plotted with a downward-pointing open triangle (shape `6` by default) at
a configurable floor value. The numeric part of the string is used as the floor
when it is lower than `bdl_floor`; otherwise `bdl_floor` is used.

The floor defaults to `0.1`, which keeps BDL points visible on a log10 axis
without distorting the range. You can change it globally or per panel:

```{r bdl-floor, eval = FALSE}
# Global floor via make_clinical_figure (not currently a direct argument —
# set it per panel using lab_panel(bdl_floor = ...))
panel_low_bdl <- lab_panel(
  line_params = "WBC (/µL)",
  bdl_floor   = 0.01
)
```

A ghost legend entry for the BDL symbol is always included in multi-case mode
so the symbol is explained even when all BDL values happen to fall off-screen.

---

## 7  Saving the figure

`save_clinical_figure()` wraps `ggplot2::ggsave()` with sensible defaults. The
output format is inferred from the file extension.

```{r save, eval = FALSE}
save_clinical_figure(fig, "figure.pdf")
save_clinical_figure(fig, "figure.png", dpi = 300)
save_clinical_figure(fig, "figure.svg", width = 14, height = 16, units = "cm")
```

| Argument | Default | Description                                 |
|----------|---------|---------------------------------------------|
| `width`  | `7`     | Figure width                                |
| `height` | `8`     | Figure height                               |
| `units`  | `"in"`  | `"in"`, `"cm"`, `"mm"`, or `"px"`          |
| `dpi`    | `300`   | Resolution (raster formats only)            |

---

## 8  Using the theme and helper layers directly

`theme_clinical()` and `named_vlines()` are exported for use in custom ggplots
outside the main workflow.

```{r theme-example, fig.height = 3}
library(ggplot2)

ggplot(data.frame(x = 1:10, y = rnorm(10)), aes(x, y)) +
  geom_line() +
  named_vlines(c("Event A" = 3, "Event B" = 7)) +
  theme_clinical(show_x = TRUE) +
  labs(x = "Day", y = "Value")
```

---

## 9  Summary of the main functions

| Function                | Purpose                                                  |
|-------------------------|----------------------------------------------------------|
| `example_data()`        | Load fictional example lab or treatment data             |
| `prep_lab_data()`       | Standardise a data frame with pre-computed relative days |
| `prep_treatment_data()` | Standardise a treatment data frame                       |
| `load_lab_data()`       | Read lab data from Excel, computing relative days        |
| `load_treatment_data()` | Read treatment data from Excel, computing relative days  |
| `lab_panel()`           | Build a panel specification object                       |
| `make_clinical_figure()`| Assemble all panels into a stacked patchwork figure      |
| `save_clinical_figure()`| Save the figure to PDF/PNG/SVG/TIFF                     |
| `theme_clinical()`      | Apply the clinical ggplot2 theme to any plot             |
| `named_vlines()`        | Add labelled vertical reference lines to any ggplot      |
